**Bayesian statistics for longitudinal studies in biomedical research**

### 2020-10-28 Manuscripts/Manuscript_AM_v1.Rmd
Updated intro check code

bibliography: refs.bib
output:
  pdf_document: 
    fig_caption: yes
  html_document: default
  css: style.css
  word_document:
    reference_docx: docx_template.docx
mainfont: Times New Roman
link-citations: yes
subtitle: '_Their application and use in biomedical research_'
biblio-style: apalike
---

@@

Researchers have typically employed a _frequentist_ approach to analyze longitudinal data. Such statistical view derives its name from the fact that it views probability as a limiting frequency [wagenmakers2008]. In practice, this analysis for longitudinal data is  based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). However, rm-ANOVA makes two key assumptions regarding the data: constant correlation across same-subject measurements and  complete observations from each subject [@schober2018;@gueorguieva2004]. The first condition is frequently unjustified in many cases as  the correlation between measures diminishes when the time gap between them increases [@ugrinowitsch2004], and in that case violating that assumption increases the type I error rate[@lane2016]. The reasons behind the second assumption being not met are varied: subjects can withdraw during the course of the study, attrition in animals due to injury or weight loss, or complications can arise that prevent the researcher from collecting measurements at a required timepoint. When incomplete observations  exist, rm-ANOVA requires to exclude the subject with missing observations from the analysis, which can result in increased costs for the study and possibly bias the results.

Moreover, rm-ANOVA uses a _post hoc_ analysis to distinguish the differences between groups. Because a _post hoc_ analysis is based in multiple repeated comparisons, it can inflate the false positivity rate[@liu2010]. Recently, _linear mixed effects models_ (LMEMs) have started to be used by certain groups to analyze longitudinal data [@vishwanath2009;@skala2010]. Generally speaking, LMEMs incorporate both _fixed_ and _random_ effects making them more flexible that rm-ANOVA as they can work with missing observations, and they allow to model the covariance of the parameters in different manners[@west2014].

However, the nature of LMEMs and rm-ANOVA restrict the inferences they can extract from a longitudinal study when the data does not follow a linear trend (Figure 1. with line and "wiggly plot"), because the model in both cases does not allow for a consistent fit with the trend of the data. This particular non-linear behavior in longitudinal data has been reported in particular in studies that measure tumor response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. In such circumstances, even if a _p-value_ with "significance" ( _p_<0.05) is obtained, the model lacks predictive power and this compromises the extent of the inferences that can be derived from the analysis.

In contrast to the _frequentist_ rm-ANOVA and LMEM analysis, _Bayesian statistics_ represent a relatively new field that does not rely on _p-values_ and hypothesis test to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the comparisons and inferences derived form the analysis.On the other hand, the shift that Bayesian theory represents from the traditional statistical view in research and the set of computational tools required for the implementation of this type of models have limited their use in the biomedical research community. Based on this, the goals of this study are: a) to present the limitations of a _frequentist_ approach (rm-ANOVA) over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies. With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.

>>>>

bibliography: refs.bib
output:
  html_document: default
  pdf_document: 
    fig_caption: yes
  css: style.css
  word_document:
    reference_docx: docx_template.docx
mainfont: Times New Roman
link-citations: yes
subtitle: '_Their application and use in biomedical research_'
biblio-style: apalike
---

@@

Researchers have typically employed a _frequentist_ approach to analyze longitudinal data, a statistical view that derives its name from the fact that it regards probability as a limiting frequency [wagenmakers2008]. when applied to longitudinal, a frequentist analysis is  based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). This methodology makes two key assumptions regarding longitudinal data: a constant correlation across same-subject measurements and  complete observations from each subject [@schober2018;@gueorguieva2004]. The first condition is frequently unjustified  as  the correlation between measures diminishes when the time gap between them increases [@ugrinowitsch2004], and in that case violating that assumption increases the type I error rate[@lane2016]. The reasons behind the second assumption being not met in biomedical research are varied: subjects can withdraw during the course of the study, attrition in animals due to injury or weight loss, or complications can arise that prevent the researcher from collecting measurements at a certain timepoint. When these issues arise, rm-ANOVA requires to exclude the subject with missing observations from the analysis, which can lead to increased costs for the study if the desired statistical power is not met with the remaining subjects, and can lead to the exclusion of valuable information.Additionaly, rm-ANOVA uses a _post hoc_ analysis to assess significance in the differences between groups. Because a _post hoc_ analysis is based in multiple repeated comparisons, it can inflate the false positivity rate[@liu2010].

Recently, _linear mixed effects models_ (LMEMs) have been used by certain groups to analyze biomedical longitudinal data [@vishwanath2009;@skala2010]. Briefly, these models incorporate _fixed_ effects, which correspond to the levels of experimental factors in the study (e.g. the different drug regimens in a clinical trial), and _random_ effects, which account for random variation within the population [@pinheiro2006]. These models are more flexible than rm-ANOVA as they can accomodate missing observations in the data, and  allow to model the covariance of the parameters in different manners[@west2014]. On the other hand, they impose restrictions in the distribution of the errors and of the random effects, and assume linearity between the response and the parameters[@schielzeth2020;@pinheiro2006].

Additionaly, the inherent nature of both LMEMs and rm-ANOVA restrict the inferences they can extract from a longitudinal study when the data does not follow a linear trend, because the model in both cases does not provide consistent fit with the trend of the data. This particular non-linear behavior is exemplified in longitudinal studies that measure tumor response to radio/chemotherapy in preclinical and clinical settings  [@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. In such circumstances, it is even possible to obtain a "significant" _p-value_( _p_<0.05) due to the well-known variability of this metric [@nuzzo2014;@halsey2019], but the model will lack predictive power and this in turn will compromise the extent of the inferences that can be derived.

In contrast to rm-ANOVA and LMEMs, _Bayesian statistics_ represent a relatively new field that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian theory represents from the traditional statistical view in research and the set of computational tools required for the implementation of this type of models have limited their use in the biomedical research community. Based on this, the goals of this study are: a) to present the limitations of a _frequentist_ approach (rm-ANOVA) over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement b) over a set of simulated data that matches previously reported trends in longitudinal biomedical studies. With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.
